---
title: "DS Capstone"
author: "Sandeep"
date: "2026-06-04"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = FALSE,
  warning    = FALSE,
  message    = FALSE,
  fig.width  = 9,
  fig.height = 5,
  fig.align  = "center"
)
library(ggplot2)
library(dplyr)
library(stringr)
library(scales)
```

## Introduction

This milestone report documents the exploratory data analysis.The goal
is to build a predictive text application, similar to smartphone
keyboard suggestions, that predicts the next word a user will type.

The three source corpora used are:

Source - Description Blogs - Personal blog entries\
News - News articles Twitter: Short-form social-media posts

# Loading the Data

```{r load-data}
blogs_path   <- "C:/Users/User/Downloads/Coursera-SwiftKey/final/en_US/en_US.blogs.txt"
news_path    <- "C:/Users/User/Downloads/Coursera-SwiftKey/final/en_US/en_US.news.txt"
twitter_path <- "C:/Users/User/Downloads/Coursera-SwiftKey/final/en_US/en_US.twitter.txt"

blogs   <- readLines(blogs_path, encoding="UTF-8", skipNul=TRUE)
news    <- readLines(news_path, encoding="UTF-8", skipNul=TRUE)
twitter <- readLines(twitter_path, encoding="UTF-8", skipNul=TRUE)

cat("Blogs lines:", length(blogs), "\n")
cat("News lines:", length(news), "\n")
cat("Twitter lines:", length(twitter), "\n")
```

# File Summary Statistics

```{r summary-stats}
file_size_mb <- function(path) {
  round(file.info(path)$size / 1e6, 1)
}

count_words <- function(lines) {
  sum(str_count(lines, "\\S+"))
}

summary_df <- data.frame(
  File    = c("Blogs", "News", "Twitter"),
  Lines   = c(length(blogs), length(news), length(twitter)),
  Words   = c(count_words(blogs), count_words(news), count_words(twitter)),
  Size_MB = c(
    file_size_mb(blogs_path),
    file_size_mb(news_path),
    file_size_mb(twitter_path)
  )
)

knitr::kable(summary_df,
             col.names = c("File", "Lines", "Words", "Size (MB)"),
             caption = "File Summary Table")
```

## Key Observations

- Twitter has the most number of lines (\~2.36 million), but each entry
  is very short due to the character limitation, making it more
  fragmented compared to other datasets.

- Blogs and News contain significantly fewer lines, but each line is
  much longer and more information-dense.

- Overall, the combined dataset is large, with a total size of nearly
  583 MB, making it suitable for building a robust text prediction
  model.

# Exploratory Data Analysis

## Words Per Line Distribution

```{r wpl-dist}
set.seed(42)
sample_size <- 5000

# Sample and calculate words per line
blogs_wpl   <- str_count(sample(blogs,   min(sample_size, length(blogs))),   "\\S+")
news_wpl    <- str_count(sample(news,    min(sample_size, length(news))),    "\\S+")
twitter_wpl <- str_count(sample(twitter, min(sample_size, length(twitter))), "\\S+")

# Combine into one data frame
wpl_df <- data.frame(
  Source = rep(c("Blogs", "News", "Twitter"),
               times = c(length(blogs_wpl), length(news_wpl), length(twitter_wpl))),
  WPL = c(blogs_wpl, news_wpl, twitter_wpl)
)

# Summary statistics
avg_wpl <- aggregate(WPL ~ Source, data = wpl_df,
                     FUN = function(x) c(Mean = mean(x), Median = median(x)))

# Clean output format
avg_wpl <- data.frame(
  Source = avg_wpl$Source,
  Mean   = round(avg_wpl$WPL[, "Mean"], 1),
  Median = avg_wpl$WPL[, "Median"]
)

knitr::kable(avg_wpl,
  caption = "Average & Median Words Per Line",
  align = c("l", "r", "r")
)
```

```{r plot-wpl, fig.cap="Figure 1 - Word-count distribution per line (capped at 150 words)"}
ggplot(wpl_df %>% filter(WPL <= 150),
       aes(x = WPL)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white", alpha = 0.8) +
  facet_wrap(~ Source) +
  labs(
    title = "Distribution of Words Per Line",
    subtitle = "Sample of 5,000 lines from each corpus (capped at 150 words)",
    x = "Words per line",
    y = "Count"
  ) +
  theme_minimal()
```

> **Finding:** Twitter lines are heavily concentrated below 20 words.
> Blogs and News follow a right-skewed distribution with longer lines
> common.

# Create sample combined dataset
```{r top-all}
sample_all <- data.frame(
  text = c(
    sample(blogs, 10000),
    sample(news, 10000),
    sample(twitter, 10000)
  ),
  Source = rep(c("Blogs", "News", "Twitter"), each = 10000)
)

library(tidytext)
library(dplyr)

top_all <- sample_all %>%
  unnest_tokens(Word, text) %>%
  count(Source, Word, sort = TRUE) %>%
  group_by(Source) %>%
  slice_max(n, n = 20)

```{r eda-plots, fig.height=9, fig.cap="EDA overview: word distribution and frequent words"}

library(gridExtra)

# Histogram: words per line
p1 <- ggplot(wpl_df %>% filter(WPL <= 120),
             aes(x = WPL)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white") +
  facet_wrap(~ Source) +
  labs(title = "Words per Line Distribution",
       x = "Words per line",
       y = "Count") +
  theme_minimal()

# Top words (FIXED: Count -> n)
p2 <- ggplot(top_all,
             aes(x = reorder(Word, n), y = n)) +
  geom_col(fill = "darkorange") +
  coord_flip() +
  facet_wrap(~ Source) +
  labs(title = "Top Words",
       x = NULL,
       y = "Frequency") +
  theme_minimal()

grid.arrange(p1, p2, nrow = 2)
```
**Key findings:**

- **Twitter sentences are shortest** — tightly clustered below 20 words, a direct result of the 140-character limit.
- **Blogs are the most verbose** — right-skewed distribution with many lines exceeding 60 words; widest spread of any corpus.
- **Stop-words dominate everywhere** — "the", "and", "to" top all three corpora; removing them will be essential before n-gram modelling.
- **"I" is unusually prominent in Twitter** — reflects the first-person, conversational register of social media.
- **"said" is a News marker** — journalism's reliance on attributed quotations pushes it into the top 20.
- **News has the richest vocabulary** — most unique words in the sample, owing to varied subject matter and formal register.
# Future Plan {#future}

## Text Prediction Algorithm

The next word prediction system will follow these steps:

**Step 1 — Data Cleaning**

- Remove non-ASCII characters, URLs, email addresses, and numbers
- Convert to lowercase
- Strip punctuation (retain sentence boundaries)
- Remove profanity (using a blocklist)
- Sample a representative subset if memory is a constraint

**Step 2 — Tokenisation**

- Split cleaned text into individual tokens (words)
- Remove stop-words for n-gram construction (optional — kept for prediction)

**Step 3 — N-gram Construction**

- Build **unigram**, **bigram**, and **trigram** frequency tables
- Store as data frames with frequency counts
- Apply **Kneser-Ney** or **Stupid Back-off** smoothing to handle unseen n-grams

**Step 4 — Prediction Logic**

- Given a user's last 1–2 typed words, look up matching trigrams
- Fall back to bigrams, then unigrams if no match found (back-off model)
- Return the top 3–5 candidate next words with probability scores

## Shiny App Design

```

    Next Word Predictor                     

    Type your text:                            
   ┌─────────────────────────────────────────┐
   │  I am going to the ...                  │
   └─────────────────────────────────────────┘
                                              
   Suggested next words:                      
   [ store ]  [ park ]  [ gym ]  [ beach ]    
                                              
   ─────────────────────────────────────────  
   Top Predictions (with confidence):         
    1. store   -           42 %               
    2. park    -           18 %               
    3. gym     -           12 %               

```

Key Shiny app features:

- **Text input box** — user types freely; prediction updates reactively
- **Word suggestion buttons** — click to append the predicted word
- **Confidence bar chart** — visualises prediction probabilities
- **Source selector** — optional filter: Blogs / News / Twitter / All
- **Fast response** — pre-computed n-gram tables loaded into memory at startup
# Conclusion

This report demonstrates successful loading and analysis of the HC Corpora datasets. The data reveals distinct writing styles across sources, particularly Twitter's brevity and Blogs' longer format. A back-off n-gram model and interactive Shiny app are planned to provide efficient next-word prediction.
